Refunctionalization of Abstract Abstract Machines (Functional Pearl)

### 2018-05-31 paper/paper.tex
update ref; fix format and code

Defining a language by building an interpreter for it can be traced to the very
early days of programming languages research \cite{landin1966next, reynolds1972definitional}.
Nowadays, even an undergraduate student in computer science is able to build toy languages
through interpreters.
But building a sound abstract interpreter remained an esoteric and difficult task until very recently.

\citeauthor{van2012systematic} proposed the Abstracting Abstract Machines (AAM) methodology which provides a recipe
for constructing sound abstract interpreters for higher-order functional languages from concrete abstract machines
~\cite{van2012systematic, van2010abstracting}.
Given a concrete small-step abstract machine (e.g. the CESK machine, Krivine's machine, etc.),
by allocating continuations in the store and bounding both the value addresses and continuation
addresses to be finite, we obtain an abstract interpreter with a finite state space which
can be used for performing sound static analysis.
One can further instantiate different polyvariant control flow analyses by using different address allocators \cite{Gilray:2016:ACP:2951913.2951936}.

@@

In this section, we derive the abstracting abstract machine from concrete
CESK machines, and also show how to instantiate
useful $k$-call-sensitive control flow analysis.

@@

In this section, we describe a variant of AAM that allows the stack to be
unbounded which uses a precise call stack as we did in the concrete CESK machine.
Instead of allocating continuations in the store and embedding
addresses of continuations in states, we intend to use a list of frames
to explicitly model the stack.
By doing so, we recover the call stack as precise as runtimes
(so called pushdown control flow analysis), but since
the stack is unbounded, the analysis is potentially not computable.
For readers who are not familiar with pushdown analysis, we have a detailed
discussion in Section~\ref{pdcfarevisit}.
The reason we show it here is to establish an artifact with precise call/return matching
used for the next step of transformation.

@@

For clarity, we first show how to refunctionalize normal continuations and
nondeterministic continuation without collecting semantics.
Then, adding an appropriate caching algorithm to make sure the analysis will
terminate at some fixed-point is a simple transformation of the program to cache-passing
style. Finally, we examine how computable pushdown control flow analysis is established
through refunctionalization and caching.

@@

\begin{lstlisting}
def analyze(e: Expr) = {
  def iter(cache: Cache): Ans = {
    val Ans(vss, new_cache) = aeval(e, Map[String, BAddr](), Store[BAddr, Storable](Map()), List(), cache, {
      val initConfig = Config(e, Map[String, BAddr](), Store[BAddr, Storable](Map()), List())
      case Ans(vss, cache) => Ans(vss, cache.outUpdate(initConfig, vss))
    })
    if (new_cache.out == cache.out) { Ans(vss, new_cache) }
    else { iter(Cache(new_cache.out, new_cache.out)) }
  }
  iter(mtCache)
}
\end{lstlisting}

@@

\subsection{Pushdown Control Flow Analysis, Revisited} \label{pdcfarevisit}

In the previous section, we have established a computable pushdown control-flow
analysis through refunctionalization and caching. In this section,
we revisit the pushdown control flow problem and examine what we have done to
overcome it.

\textbf{The Problem with Return-flows.}
Pushdown control flow is a property in analysis that precisely models
the runtime call stack of the analyzed program.
A pushdown control flow analysis provides an as-exact-as-runtime return-flow
when analyzing a program, but traditional control flow analysis collapses
the state space into a finite space and thus causes imprecise stack modeling.

To see how traditional control flow analysis suffers from spurious return-flows,
we can consider the following example:

@@

In $k$-CFA algorithm or the abstract abstract machine shown in Section~\ref{aam},
the call sites @(id 2)@ and @(id 1)@ share the same return flow,
so the invocation of @(id 2)@ returns to both call-sites @[x (id 1)]@ and @[y (id 2)]@.
Therefore, the returned value @2@ for variable @y@ is also propagated to
variable @x@, causing imprecise analysis results to arise.
This return-flow merging is inevitable, even when increasing the context-sensitivity.
If we use the monovariant analysis (i.e., $0$-CFA), the analysis result would be such that @x@ and @y@ point to value set @{1, 2}@, because the algorithm does not distinguish that
@z@ comes from a different call site.
Under $1$-CFA, the algorithm is able to distinguish that variable @z@ of function
@id@ has two different values at two call sites, so variable @y@
would not be polluted by @1@.
However, variable @x@ still points to value set @{1, 2}@, because two call-sites still share the same continuation $1$-CFA is still unable to separate the return-flows.

@@

We have multiple choices of how to proceed:
\begin{itemize}
  \item By monadifying the continuations, we obtain the abstracting
    definitional interpreters in monadic style which are what \citeauthor{darais2017abstracting}
    describe \cite{darais2017abstracting}.
  \item By representing the extended continuation-passing style with delimited controls,
    we derive a new form of abstract interpreters in direct-style.
  \item In fact, we may also just use the same caching algorithm but with side effects such as
    assignments and mutations to update the cache, and then achieve the same
    definitional interpreter with pushdown control flows. In this case, nondeterminism can be
    easily handled via @for@ comprehension.
\end{itemize}

@@

\begin{lstlisting}
val mtEnv = Map[String, BAddr]()
val mtStore = Store[BAddr, Storable](Map())
val mtTime = List()

def analyze(e: Expr) = {
  def iter(cache: Cache): Ans = {
    val Ans(vss, anscache) = reset { aeval(e, mtEnv, mtStore, mtTime, cache) }
    val initConfig = Config(e, Map[String, BAddr](), Store[BAddr, Storable](Map()), List())
    val new_cache = anscache.outUpdate(initConfig, vss)
    if (new_cache.out == cache.out) { Ans(vss, new_cache) }
    else { iter(Cache(new_cache.out, new_cache.out)) }
  }
  iter(mtCache)
}
\end{lstlisting}

@@

\textit{Abstract Interpretation and Control-flow Analysis.}
\citeauthor{cousot1977abstract} invented abstract interpretation as a sound approach
to approximate a program's runtime behavior\cite{cousot1977abstract}.
Control flow analysis is one instance of abstract interpretation on functional programs
that can be traced to Jones \cite{jones1981flow}.
\citeauthor{Shivers:1988:CFA:53990.54007} introduces $k$-CFA which uses $k$ recent calling contexts as program contour that differentiates values from different contexts \cite{Shivers:1988:CFA:53990.54007, Shivers:1991:SSC:115865.115884}.
The modern formulation of control flow analysis is the abstracting abstract machines
methodology \cite{van2012systematic, van2010abstracting} which forms the starting point of this paper.
Modern programming languages such as Java\cite{might2010resolving} and Racket \cite{Tobin-Hochstadt:2012:HSE:2384616.2384655}
can also be modeled by the abstracting abstract machine approach. \\

\textit{Defunctionalization and Refunctionalization.}
Defunctionalization and refunctionalization build connections between abstract
machines and interpreters.
\citeauthor{reynolds1972definitional} first shows defunctionalization as a program transformation technique that
can be used to transform higher-order functions to first-order functions
when specifying a programming language by an interpreter \cite{reynolds1972definitional}.
Refunctionalization is the reverse of this transformation.

@@

\textit{Pushdown Control Flow Analysis.}
In recent years, there are significant efforts~\cite{vardoulakis2010cfa2, earl2012introspective,
gilray2016pushdown, johnson2015abstracting} to achieve precise call/return
match based on small-step abstracting abstract machines.

CFA2 is the first solution that solves the return-flows problem~\cite{vardoulakis2010cfa2},
but CFA2 has several drawbacks: it works only on continuation-passing style programs,
and does not supports polyvariant analysis, in addition to an exponential time
complexity.
Pushdown control flow analysis (PDCFA) is a mechanism which maintains this precision
through the use of a Dyke state graph representing all possible stacks contained
within the unbounded-stack machine\cite{earl2012introspective, earl2010pushdown}.
Similar to PDCFA, abstracting abstract control (AAC) is another strategy for
maintaining stack precision \cite{johnson2015abstracting}.
AAC functions by utilizing continuations which are specific to both
the source and target states of a call-site transition,
which guarantees that no spurious merging will occur during returns.

@@

\textit{Abstracting Definitional Interpreters.}
Reynold's seminal paper \emph{Definitional interpreters for higher-order
  programming languages} \cite{reynolds1972definitional} shows that in
definitional interpreters, the defined language can inherit properties from the defining language.
With this insight, \citeauthor{darais2017abstracting} construct abstracting
definitional interpreters which automatically inherit the pushdown control-flow
property from its defining language because the defined language simply uses the stack of the
meta-language~\cite{darais2017abstracting}. \citeauthor{darais2017abstracting}'s abstracting
definitional interpreters work on direct-style programs, and is written in monadic style. One
of the advantages of monadic abstract interpreters is modular and composable, therefore
deploying different sensitivities or features is just applying a different monad.
Prior to that, \citeauthor{Sergey:2013:MAI:2491956.2491979} show a monadic
abstract interpreter for small-step semantics \cite{Sergey:2013:MAI:2491956.2491979}.

>>>>

Defining a language by building an interpreter for it can be traced to the very
early days of programming languages research \cite{landin1966next, Reynolds:72}.
Nowadays, even an undergraduate student in computer science is able to build toy languages
through interpreters.
But building a sound abstract interpreter remained an esoteric and difficult task until very recently.

\citeauthor{van2012systematic} proposed the Abstracting Abstract Machines (AAM) methodology which provides a recipe
for constructing sound abstract interpreters for higher-order functional languages from concrete abstract machines
~\cite{van2012systematic, van2010abstracting}.
Given a concrete small-step abstract machine (e.g. the CESK machine, Krivine's machine, etc.),
by allocating continuations in the store and bounding both the value addresses and continuation
addresses to be finite, we obtain an abstract interpreter with a finite state space which
can be used for performing sound static analysis.
One can further instantiate different polyvariant control-flow analyses by using different address allocators \cite{Gilray:2016:ACP:2951913.2951936}.

@@

In this section, we derive the abstracting abstract machine from concrete
CESK machines, and also show how to instantiate
useful $k$-call-sensitive control-flow analysis.

@@

In this section, we describe a variant of AAM that allows the stack to be
unbounded which uses a precise call stack as we did in the concrete CESK machine.
Instead of allocating continuations in the store and embedding
addresses of continuations in states, we intend to use a list of frames
to explicitly model the stack.
By doing so, we recover the call stack as precise as runtimes
(so called pushdown control-flow analysis), but since
the stack is unbounded, the analysis is potentially not computable.
For readers who are not familiar with pushdown analysis, we have a detailed
discussion in Section~\ref{pdcfarevisit}.
The reason we show it here is to establish an artifact with precise call/return matching
used for the next step of transformation.

@@

For clarity, we first show how to refunctionalize normal continuations and
nondeterministic continuation without collecting semantics.
Then, adding an appropriate caching algorithm to make sure the analysis will
terminate at some fixed-point is a simple transformation of the program to cache-passing
style. Finally, we examine how computable pushdown control-flow analysis is established
through refunctionalization and caching.

@@

\begin{lstlisting}
val mtEnv = Map[String, BAddr](); val mtStore = Store[BAddr, Storable](Map()); val mtTime = List()
val mtCache = Cache(Store[Config, VS](Map()), Store[Config, VS](Map()))

def analyze(e: Expr) = {
  def iter(cache: Cache): Ans = {
    val Ans(vss, new_cache) = aeval(e, mtEnv, mtStore, mtTime, cache, {
      case Ans(vss, cache) => 
        val initConfig = Config(e, mtEnv, mtStore, mtTime) 
        Ans(vss, cache.outUpdate(initConfig, vss))
    })
    if (new_cache.out == cache.out) { Ans(vss, new_cache) }
    else { iter(Cache(new_cache.out, new_cache.out)) }
  }
  iter(mtCache)
}
\end{lstlisting}

@@

\subsection{Pushdown Control-Flow Analysis, Revisited} \label{pdcfarevisit}

In the previous section, we have established a computable pushdown control-flow
analysis through refunctionalization and caching. In this section,
we revisit the pushdown control-flow problem and examine what we have done to
overcome it.

\textbf{The Problem with Return-flows.}
Pushdown control-flow is a property in analysis that precisely models
the runtime call stack of the analyzed program.
A pushdown control-flow analysis provides an as-exact-as-runtime return-flow
when analyzing a program, but traditional control-flow analysis collapses
the state space into a finite space and thus causes imprecise stack modeling.

To see how traditional control-flow analysis suffers from spurious return-flows,
we can consider the following example:

@@

In $k$-CFA algorithm or the abstract abstract machine shown in Section~\ref{aam},
the call sites @(id 2)@ and @(id 1)@ share the same return flow,
so the invocation of @(id 2)@ returns to both call-sites @[x (id 1)]@ and @[y (id 2)]@.
Therefore, the returned value @2@ for variable @y@ is also propagated to
variable @x@, causing imprecise analysis results to arise.
This return-flow merging is inevitable, even when increasing the context-sensitivity.
If we use the monovariant analysis (i.e., $0$-CFA), the analysis result would be such 
that @x@ and @y@ point to value set @{1, 2}@, because the algorithm does not distinguish 
that @z@ comes from a different call site.
Under $1$-CFA, the algorithm is able to distinguish that variable @z@ of function
@id@ has two different values at two call sites, so variable @y@
would not be polluted by @1@.
However, variable @x@ still points to value set @{1, 2}@, because two call-sites still 
share the same continuation and $1$-CFA is still unable to separate the return-flows.

@@

We have multiple choices of how to proceed:
\begin{itemize}
  \item By monadifying the continuations, we obtain the abstracting
    definitional interpreters in monadic style which are what \citeauthor{darais2017abstracting}
    describe \cite{darais2017abstracting}.
  \item By representing the extended continuation-passing style with delimited controls,
    we derive a new form of abstract interpreters in direct-style.
  \item In fact, we may also just use the same caching algorithm but with side effects such as
    assignments and mutations to update the cache, and then achieve the same
    definitional interpreter with pushdown control-flows. In this case, nondeterminism can be
    easily handled via @for@ comprehension.
\end{itemize}

@@

\begin{lstlisting}
def analyze(e: Expr) = {
  def iter(cache: Cache): Ans = {
    val Ans(vss, anscache) = reset { aeval(e, mtEnv, mtStore, mtTime, cache) }
    val initConfig = Config(e, mtEnv, mtStore, mtTime) 
    val new_cache = anscache.outUpdate(initConfig, vss)
    if (new_cache.out == cache.out) { Ans(vss, new_cache) }
    else { iter(Cache(new_cache.out, new_cache.out)) }
  }
  iter(mtCache)
}
\end{lstlisting}

@@

\textit{Abstract Interpretation and Control-Flow Analysis.}
\citeauthor{cousot1977abstract} invented abstract interpretation as a sound approach
to approximate a program's runtime behavior\cite{cousot1977abstract}.
Control-flow analysis is one instance of abstract interpretation on functional programs
that can be traced to Jones \cite{jones1981flow}.
\citeauthor{Shivers:1988:CFA:53990.54007} introduces $k$-CFA which uses $k$ recent calling contexts as program contour that differentiates values from different contexts \cite{Shivers:1988:CFA:53990.54007, Shivers:1991:SSC:115865.115884}.
The modern formulation of control-flow analysis is the abstracting abstract machines
methodology \cite{van2012systematic, van2010abstracting} which forms the starting point of this paper.
Modern programming languages such as Java\cite{might2010resolving} and Racket \cite{Tobin-Hochstadt:2012:HSE:2384616.2384655}
can also be modeled by the abstracting abstract machine approach. \\

\textit{Defunctionalization and Refunctionalization.}
Defunctionalization and refunctionalization build connections between abstract
machines and interpreters.
\citeauthor{Reynolds:72} first shows defunctionalization as a program transformation technique that
can be used to transform higher-order functions to first-order functions
when specifying a programming language by an interpreter \cite{Reynolds:72, Reynolds:HOSC98-revisited}.
Refunctionalization is the reverse of this transformation.

@@

\textit{Pushdown Control-Flow Analysis.}
In recent years, there are significant efforts~\cite{vardoulakis2010cfa2, earl2012introspective,
gilray2016pushdown, johnson2015abstracting} to achieve precise call/return
match based on small-step abstracting abstract machines.

CFA2 is the first solution that solves the return-flows problem~\cite{vardoulakis2010cfa2},
but CFA2 has several drawbacks: it works only on continuation-passing style programs,
and does not supports polyvariant analysis, in addition to an exponential time
complexity.
Pushdown control-flow analysis (PDCFA) is a mechanism which maintains this precision
through the use of a Dyke state graph representing all possible stacks contained
within the unbounded-stack machine\cite{earl2012introspective, earl2010pushdown}.
Similar to PDCFA, abstracting abstract control (AAC) is another strategy for
maintaining stack precision \cite{johnson2015abstracting}.
AAC functions by utilizing continuations which are specific to both
the source and target states of a call-site transition,
which guarantees that no spurious merging will occur during returns.

@@

\textit{Abstracting Definitional Interpreters.}
Reynold's seminal paper \emph{Definitional interpreters for higher-order programming languages} 
\cite{Reynolds:72, Reynolds:HOSC98-revisited} shows that in
definitional interpreters, the defined language can inherit properties from the defining language.
With this insight, \citeauthor{darais2017abstracting} construct abstracting
definitional interpreters which automatically inherit the pushdown control-flow
property from its defining language because the defined language simply uses the stack of the
meta-language~\cite{darais2017abstracting}. \citeauthor{darais2017abstracting}'s abstracting
definitional interpreters work on direct-style programs, and is written in monadic style. One
of the advantages of monadic abstract interpreters is modular and composable, therefore
deploying different sensitivities or features is just applying a different monad.
Prior to that, \citeauthor{Sergey:2013:MAI:2491956.2491979} show a monadic
abstract interpreter for small-step semantics \cite{Sergey:2013:MAI:2491956.2491979}.
